---
title: "The {uncounted} package"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics

The {uncounted} package implements various methods to estimate population size such as: graph-based methods, non-linear (count) regression models or dual/multiple system estimation.

To install the package from GitHub you can use the `pak` package 

```{r, eval=FALSE}
install.packages("pak")
pak::pkg_install("ncn-foreigners/uncounted")
```

Load the package

```{r}
library(uncounted)
```

Below you can find the basic functionality of the package on the Polish example from 2018

```{r}
data(foreigners_pl)
model_data <- subset(foreigners_pl, year == 2018 & half == 1)
head(model_data)
```

Fit the basic Zhang model as described [here](https://www.ssb.no/en/befolkning/artikler-og-publikasjoner/developing-methods-for-determining-the-number-of-unauthorized-foreigners-in-norway--18906) to obtain estimate of the irregular population size along with 95% confidence interval.

```{r}
result_zhang <- estimate_hidden_pop(
   data = model_data,
   observed = ~ border,
   auxiliary = ~ police,
   reference_pop = ~ pesel,
   method = "mle",
   family = "poisson"
)

c(N_hat=result_zhang$estimates$xi, 
  lower=unname(result_zhang$confint_xi[1]),
  upper=unname(result_zhang$confint_xi[2]))
```



## Acknowledgements

+ Work on this package is supported by the National Science Centre, OPUS 20 grant no. 2020/39/B/HS4/00941 (Towards census-like statistics for foreign-born populations -- quality, data integration and estimation)

+ Name of the package was motivated by the name of the [Zult, D. B. (2024). Counting the uncounted: Methodological extensions in multiple systems estimation. Dissertation, Maastricht University, doi:10.33540/2611](https://www.cbs.nl/en-gb/background/2024/51/counting-the-uncounted)

