---
title: "Model Diagnostics and Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Diagnostics and Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the diagnostic and model-comparison tools
available in the **uncounted** package. We cover goodness-of-fit
statistics, residual analysis, rootograms, exploratory log-log plots,
leave-one-out sensitivity analysis, and gamma profiling. Together these
tools help you decide which model specification is most appropriate and
whether the population-size estimates are robust.

## 1. Setup: fitting Poisson and Negative Binomial models

We use the `irregular_migration` dataset shipped with the package. Two
models are fitted---Poisson and Negative Binomial---with year and sex
covariates on alpha and year covariates on beta.

```{r fit-models}
library(uncounted)
data(irregular_migration)

fit_po <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "poisson",
  cov_alpha = ~ factor(year) + sex, cov_beta = ~ factor(year),
  countries = ~ country_code
)

fit_nb <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "nb",
  cov_alpha = ~ factor(year) + sex, cov_beta = ~ factor(year),
  countries = ~ country_code
)
```

Quick summaries:

```{r summaries}
summary(fit_po)
summary(fit_nb)
```

## 2. Model comparison

### Comparison table

`compare_models()` produces a side-by-side table of log-likelihood, AIC,
BIC, deviance, Pearson chi-squared, RMSE, and three pseudo $R^2$ measures:

- **`R2_cor`**: $\mathrm{cor}(m, \hat\mu)^2$ --- squared correlation between
  observed and fitted values.
- **`R2_D`**: Explained deviance $1 - D(\text{model}) / D(\text{null})$, where
  the null model is the same specification without covariates (single $\alpha$,
  single $\beta$). Measures how much covariates improve the fit beyond the
  baseline power-law structure.
- **`R2_CW`**: Cameron--Windmeijer (1996) pseudo $R^2$, which uses the
  model-implied variance function $V(\mu)$ --- specifically designed for count
  data regression.

Models are sorted by AIC (the default). Note that `R2_D` compares the fitted
model against a null model (same method, no covariates). If the model has no
`cov_alpha` / `cov_beta`, then `R2_D = 0` by construction --- it only becomes
informative when covariates are present.

```{r compare-models}
comp <- compare_models(Poisson = fit_po, NB = fit_nb, sort_by = "AIC")
comp
```

Lower AIC and BIC values indicate a better trade-off between fit and
complexity. The Pearson chi-squared statistic divided by its degrees of
freedom (approximately `n_obs - n_par`) gives a quick dispersion check:
values much greater than 1 suggest overdispersion, which favours the NB
model over Poisson.

### Likelihood ratio test

Because the Poisson model is nested within the Negative Binomial (the
Poisson is the NB with theta -> infinity), we can use a likelihood ratio
test. `lrtest()` applies the boundary correction of Self & Liang (1987)
automatically when comparing Poisson against NB, because the dispersion
parameter lies on the boundary of the parameter space under H0.

```{r lrtest}
lrtest(fit_po, fit_nb)
```

A small p-value indicates significant overdispersion, favouring the NB
model.

## 3. Residual diagnostics

### Residual types

The `residuals()` method supports four types:

- **response**: raw residuals, $m_i - \hat\mu_i$.
- **pearson**: standardised by the variance function, $(m_i - \hat\mu_i) / \sqrt{V(\hat\mu_i)}$.
- **deviance**: signed square root of individual deviance contributions.
- **anscombe**: variance-stabilising residuals (approximately standard normal for a well-specified model).

```{r residuals}
r_resp <- residuals(fit_nb, type = "response")
r_pear <- residuals(fit_nb, type = "pearson")
r_dev  <- residuals(fit_nb, type = "deviance")
r_ansc <- residuals(fit_nb, type = "anscombe")

summary(r_pear)
summary(r_ansc)
```

### Four-panel diagnostic plot

`plot()` on a fitted model produces the Zhang (2008) four-panel display:

1. **Fitted vs Observed (sqrt scale)** -- points should cluster around the 45-degree line.
2. **Anscombe residuals vs fitted** -- the running mean should be flat near zero (no systematic misfit).
3. **Scale-Location** -- absolute Anscombe residuals vs fitted; an increasing trend signals under-modelled variance.
4. **Normal Q-Q** -- Anscombe residuals should follow the reference line if the distributional assumption holds.

```{r plot-diagnostics, fig.height = 8}
op <- par(mfrow = c(2, 2))
plot(fit_nb, ask = FALSE)
par(op)
```

**What to look for:**

- Curvature in Panel 1 suggests the mean function is misspecified.
- A non-flat smoother in Panel 2 indicates systematic bias at certain fitted-value ranges.
- A fan shape in Panel 3 points to overdispersion or heteroscedasticity.
- Heavy tails or S-shapes in Panel 4 indicate departure from the assumed distribution.

Compare with the Poisson fit to see whether switching to NB resolves any
patterns:

```{r plot-diagnostics-po, fig.height = 8}
op <- par(mfrow = c(2, 2))
plot(fit_po, ask = FALSE)
par(op)
```

## 4. Rootograms

Rootograms (Kleiber & Zeileis, 2016) compare the observed and fitted
count-frequency distributions directly. Three display styles are
available:

- **hanging** (default): observed bars are hung from the fitted curve.
  If the model fits well, the bottom of each bar touches zero.
- **standing**: observed bars stand on the x-axis with the fitted curve
  overlaid. Harder to judge discrepancies visually.
- **suspended**: plots the difference $\sqrt{f_\text{obs}} - \sqrt{f_\text{exp}}$
  directly. Bars crossing zero indicate over- or under-prediction at
  that count.

```{r rootogram-nb, fig.height = 4}
rootogram(fit_nb, style = "hanging")
```

```{r rootogram-po, fig.height = 4}
rootogram(fit_po, style = "hanging")
```

Compare the two: if the Poisson rootogram shows bars consistently
hanging below the zero line at small counts and above at moderate
counts, this is a classic overdispersion signature that the NB model
should correct.

```{r rootogram-styles, fig.height = 4}
rootogram(fit_nb, style = "standing")
rootogram(fit_nb, style = "suspended")
```

## 5. Exploratory log-log plots

Before or after fitting, `plot_explore()` produces two log-log scatter
plots that reveal the marginal relationships underlying the power-law
model:

1. **log(m/N) vs log(N)**: the slope approximates $\alpha - 1$. A slope
   near zero means the apprehension rate is roughly independent of
   population size ($\alpha \approx 1$). A negative slope indicates
   decreasing returns to scale ($\alpha < 1$).
2. **log(m/N) vs log(n/N)**: the slope approximates $\beta$. A positive
   slope confirms that the auxiliary rate is predictive of the
   apprehension rate.

Observations with $m = 0$ or $n = 0$ are excluded (undefined on the log
scale).

```{r plot-explore, fig.width = 9, fig.height = 4.5}
plot_explore(fit_nb)
```

These plots are exploratory visual checks. The OLS lines overlaid are
rough guides---the formal model accounts for covariates and the gamma
offset.

## 6. Leave-one-out sensitivity analysis

LOO refits the model repeatedly, dropping one observation (or one
country) at a time, to assess how much each unit influences the total
population-size estimate.

### LOO by country

Dropping all rows for a country measures its structural contribution.
This is computationally intensive (one refit per country).

```{r loo-country, eval = FALSE}
loo_po_ctry <- loo(fit_po, by = "country", verbose = TRUE)
loo_nb_ctry <- loo(fit_nb, by = "country", verbose = TRUE)
```

```{r loo-country-print, eval = FALSE}
print(loo_nb_ctry)
summary(loo_nb_ctry)
```

The `print()` method shows the top 10 most influential countries ranked
by $|\Delta\xi|$, and `summary()` reports coefficient and xi stability
across all LOO refits.

```{r loo-country-plot, eval = FALSE}
plot(loo_nb_ctry, type = "xi")
plot(loo_nb_ctry, type = "coef")
```

The xi bar plot shows which countries pull the total estimate up (red
bars, dropping the country decreases the estimate) or down (blue bars).

### LOO by observation

Dropping individual rows identifies specific data points that drive the
estimate. Useful for outlier detection.

```{r loo-obs, eval = FALSE}
loo_nb_obs <- loo(fit_nb, by = "obs", verbose = TRUE)
print(loo_nb_obs)
plot(loo_nb_obs)
```

### Comparing LOO across models

`compare_loo()` brings together LOO results from two models and produces
a scatter plot where each axis shows the percentage change in xi when
dropping a given unit:

```{r compare-loo, eval = FALSE}
comp_loo <- compare_loo(
  loo_po_ctry, loo_nb_ctry,
  labels = c("Poisson", "NB")
)
print(comp_loo)
```

**Scatter plot interpretation (4 quadrants):**

```{r compare-loo-scatter, eval = FALSE}
plot(comp_loo, type = "scatter")
```

- **Top-right (+, +):** dropping this unit increases the estimate under
  both models. The unit was pulling the estimate down in both.
- **Bottom-left (-, -):** dropping this unit decreases the estimate
  under both models. The unit was pulling the estimate up in both.
- **Off-diagonal (top-left or bottom-right):** divergent influence. The
  unit affects the two models in opposite directions---a sign that model
  choice matters for that unit.
- **Near the origin:** non-influential under both models.

Points near the diagonal indicate consistent influence; points far from
the diagonal warrant investigation.

```{r compare-loo-bar, eval = FALSE}
plot(comp_loo, type = "bar")
```

The side-by-side bar plot ranks countries by maximum absolute percentage
change across both models, making it easy to spot the most consequential
units.

## 7. Gamma profile

`profile_gamma()` refits the model across a grid of fixed gamma values
and plots two curves:

1. **$\hat\xi(\gamma)$**: how the total population-size estimate changes
   with gamma.
2. **$\ell(\gamma)$**: the profile log-likelihood as a function of gamma.

```{r profile-gamma, fig.width = 9, fig.height = 4.5}
prof <- profile_gamma(fit_nb, gamma_grid = seq(1e-4, 0.5, length.out = 10))
```

**Interpretation:**

- A **sharp peak** in the log-likelihood curve means gamma is
  well-identified by the data. The MLE is precise and the
  population-size estimate is robust to small perturbations in gamma.
- A **flat** log-likelihood curve means gamma is weakly identified. The
  data cannot distinguish between a range of gamma values, and the
  population-size estimate may be sensitive to the gamma assumption.
  In this case, reporting results for several gamma values (or fixing
  gamma based on external information) is advisable.

If the original model estimated gamma, its point estimate is marked with
a dashed red vertical line. The xi profile shows whether the
population-size estimate is stable (flat) or sensitive (steep) across
the gamma range.

The profiling results are returned invisibly as a data frame:

```{r profile-gamma-table}
prof <- profile_gamma(fit_nb, plot = FALSE)
head(prof)
```

## 8. Dependence and threshold diagnostics

The factorization
\[
E(m_i \mid N_i, n_i) = \xi_i \rho_i
\]
is an identifying assumption, not something the observed aggregated data
can test directly. The package therefore keeps four related diagnostics
separate:

- `dependence_bounds()`: fixed-center identification envelope.
- `profile_dependence()`: model-based moving-`\xi` profile.
- `robustness_dependence()`: smallest dependence strength needed to cross a target.
- `exceedance_popsize()`: empirical bootstrap tail area for threshold questions.

### 8.1 Fixed-center identification bounds

Under the bounded sensitivity model,
\[
\mu_i = \xi_i \rho_i \kappa_i,
\qquad
1 / \Gamma \le \kappa_i \le \Gamma,
\]
the case `Gamma = 1` reproduces the baseline estimate and larger values of
`Gamma` widen the admissible range for the total `xi`.

```{r sensitivity-dependence}
sens <- dependence_bounds(fit_nb, Gamma = c(1, 1.1, 1.25, 1.5))
sens
```

The table keeps the baseline estimate fixed and reports lower and upper
bounds under progressively stronger departures from the separability
assumption. These bounds are useful when the substantive question is
whether the headline estimate would remain above or below a policy-relevant
threshold under moderate departures from independence.

### 8.2 Moving-`\xi` dependence profile

`profile_dependence()` instead treats dependence as a parametric offset
\[
\mu_i(\delta) = \exp(\delta)\,\xi_i \rho_i,
\]
refits the model for each fixed `delta`, and reports the resulting
plug-in total \(\hat\xi(\delta)\).

```{r profile-dependence}
prof_dep <- profile_dependence(
  fit_nb,
  delta_grid = seq(-0.4, 0.4, length.out = 7),
  plot = FALSE
)
prof_dep
```

Unlike `dependence_bounds()`, the estimate is allowed to move because the
model is re-fit at each grid point. This makes the profile more
model-dependent, but it answers a different question: how much does the
headline estimate shift under a structured departure from separability?

### 8.3 Robustness value for a target crossing

The profile can be summarised by the smallest absolute `delta` needed to
push the total estimate across a threshold or relative target.

```{r robustness-dependence}
robustness_dependence(
  prof_dep,
  threshold = attr(popsize(fit_nb, total = TRUE), "total")$estimate * 0.9,
  direction = "decrease"
)
```

This is the one-dimensional analogue of a robustness value: the smaller the
reported `rv_Gamma`, the less dependence is needed to overturn the chosen
substantive conclusion.

### 8.4 Bootstrap threshold probabilities

When the question is explicitly threshold-based, the bootstrap can be
summarised as an empirical tail area rather than an interval.

```{r exceedance-popsize, eval = FALSE}
boot_dep <- bootstrap_popsize(
  fit_nb,
  R = 99,
  cluster = ~ country_code,
  total = TRUE,
  verbose = FALSE
)

exceedance_popsize(boot_dep, threshold = 500000)
```

`exceedance_popsize()` reports a bootstrap exceedance probability, not a
Bayesian posterior probability. It is most useful for statements such as
"what fraction of bootstrap replications exceed 500,000?" rather than for
constructing another confidence interval.

### 8.5 Omitted-frailty sensitivity for `Xi`

The previous diagnostics either keep the baseline estimate fixed
(`dependence_bounds()`), move it through a parametric dependence offset
(`profile_dependence()`), or summarize the bootstrap distribution
(`exceedance_popsize()`). A different question is whether an omitted shared
frailty could jointly affect hidden stock and detection in a way that
changes the grouped hidden-population estimate itself.

`frailty_sensitivity()` addresses that question by linearizing the fitted
Poisson/NB mean model with its IRLS working response, indexing the omitted
frailty by weighted partial-\(R^2\) values, and mapping the resulting
first-order bias into the target \(\hat\Xi_t\).

```{r frailty-sensitivity, eval = FALSE}
frailty_sensitivity(
  fit_nb,
  by = ~ year,
  r2_d = c(0, 0.05, 0.10),
  r2_y = c(0, 0.05, 0.10),
  plot = FALSE
)
```

The output reports a grouped sensitivity surface for \(\Xi_t\) together
with robustness values for threshold or percentage-change questions. The
full derivation is documented in `?frailty_sensitivity`.

## References

- Beręsewicz, M., & Pawlukiewicz, K. (2020). Estimation of the number of irregular foreigners in Poland using non-linear count regression models. arXiv preprint arXiv:2008.09407.
- Kleiber, C. and Zeileis, A. (2016). Visualizing Count Data
  Regressions Using Rootograms. *The American Statistician*, 70(3),
  296--303.
- McCullagh, P. and Nelder, J. A. (1989). *Generalized Linear Models*,
  2nd ed. Chapman & Hall.
- Self, S. G. and Liang, K.-Y. (1987). Asymptotic properties of maximum
  likelihood estimators and likelihood ratio tests under nonstandard
  conditions. *JASA*, 82(398), 605--610.
- Zhang, L.-C. (2008). *Developing methods for determining the number of unauthorized foreigners in Norway* (Documents 2008/11). Statistics Norway. https://www.ssb.no/a/english/publikasjoner/pdf/doc_200811_en/doc_200811_en.pdf
- Beręsewicz, M., & Pawlukiewicz, K. (2020). Estimation of the number of irregular foreigners in Poland using non-linear count regression models. arXiv preprint arXiv:2008.09407.
