---
title: "Hypothesis Tests for Population Size"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis Tests for Population Size}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Goal

The `hypotheses_popsize()` function tests statements about estimated
population sizes, denoted here by $\xi$. It is designed for questions such as:

- Is the estimated population size in 2024 larger than 200,000?
- Did the estimated population size increase between 2019 and 2024?
- Is a custom contrast of several grouped population-size estimates equal to
  zero?

The function uses an explicit quantity-indexed syntax:

```r
xi[year == 2024]
xi[year == 2024] - xi[year == 2019]
```

This avoids ambiguous bare labels and leaves room for future extensions to
other quantities, such as `alpha[...]`.

```{r load-package}
library(uncounted)
data(irregular_migration)
```

## Fit A Model

For a compact example, fit a Poisson model with a fixed gamma offset and then
compute population size by year.

```{r fit}
fit <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method = "poisson",
  gamma = 0.005,
  countries = ~ country
)

ps_year <- popsize(fit, by = ~ year)
ps_year
```

The examples below use the bias-corrected estimate explicitly. The
`estimate` argument is required so that the tested estimand is never chosen
silently.

By default, directional character hypotheses are interpreted as the
alternative hypothesis. For example, `"xi[year == 2024] > 200000"` tests

\[
H_0: \xi_{2024} \le 200000
\quad\text{against}\quad
H_1: \xi_{2024} > 200000.
\]

The returned table includes explicit `null_hypothesis` and
`alternative_hypothesis` columns. If you want the character expression to be
the null hypothesis instead, use `hypothesis_side = "null"`.

## Threshold Tests

To test whether the 2024 population size is larger than 200,000:

```{r threshold}
hypotheses_popsize(
  ps_year,
  "xi[year == 2024] > 200000",
  estimate = "estimate_bc"
)
```

This is a one-sided Wald test. The contrast is

\[
\hat\xi_{2024} - 200000.
\]

A positive statistic supports the alternative that $\xi_{2024}$ is larger than
the threshold.

To treat the same directional statement as the null hypothesis, write:

```{r threshold-null}
hypotheses_popsize(
  ps_year,
  "xi[year == 2024] > 200000",
  estimate = "estimate_bc",
  hypothesis_side = "null"
)
```

This asks whether there is evidence against
$H_0: \xi_{2024} \ge 200000$ in favor of
$H_1: \xi_{2024} < 200000$.

The compact form also works when there is exactly one grouping variable:

```{r compact}
hypotheses_popsize(
  ps_year,
  "xi[2024] > 200000",
  estimate = "estimate_bc"
)
```

## Change Over Time

To test whether the estimated population size increased between 2019 and 2024:

```{r increase}
hypotheses_popsize(
  ps_year,
  "xi[year == 2024] - xi[year == 2019] > 0",
  estimate = "estimate_bc"
)
```

The standard error uses the full covariance matrix of the two estimates. This
matters because year-specific estimates are functions of the same fitted model
parameters and are generally dependent.

Two-sided equality tests use `=` or `==`:

```{r equality}
hypotheses_popsize(
  ps_year,
  "xi[year == 2024] = xi[year == 2019]",
  estimate = "estimate_bc"
)
```

## Multiple Groups

Filters can match more than one row. In that case, `xi[...]` is the sum of the
matching population-size estimates, with covariance handled before testing.

```{r multi-row}
ps_year_sex <- popsize(fit, by = ~ year + sex)

hypotheses_popsize(
  ps_year_sex,
  "xi[year == 2024] - xi[year == 2019] > 0",
  estimate = "estimate_bc"
)
```

Custom functions are useful when the contrast is easier to write in R code.
The function receives a named vector of xi estimates and, optionally, the group
metadata.

```{r function}
hypotheses_popsize(
  ps_year,
  function(x, groups) {
    c(change_2024_2019 = unname(x["2024"] - x["2019"]))
  },
  estimate = "estimate_bc"
)
```

## Statistical Theory

For analytical `popsize()` results, `hypotheses_popsize()` uses Wald inference
for nonlinear functions of fitted model parameters.

Let $\theta$ denote the fitted alpha coefficients. Under standard large-sample
regularity conditions,

\[
\hat\theta \approx N(\theta, \widehat V_\theta).
\]

Population size is a nonlinear function of those parameters:

\[
\xi = h(\theta) = \sum_i N_i^{\alpha_i}.
\]

The delta method gives the approximate sampling variance:

\[
\widehat{\mathrm{Var}}(\hat\xi)
  =
  \nabla h(\hat\theta)'
  \widehat V_\theta
  \nabla h(\hat\theta).
\]

For a threshold hypothesis such as

\[
H_0: \xi_{2024} \le c
\quad\text{against}\quad
H_1: \xi_{2024} > c,
\]

the reported statistic is

\[
z = \frac{\hat\xi_{2024} - c}{SE(\hat\xi_{2024})}.
\]

For an increase hypothesis,

\[
H_0: \xi_{2024} - \xi_{2019} \le 0
\quad\text{against}\quad
H_1: \xi_{2024} - \xi_{2019} > 0,
\]

the statistic is

\[
z =
\frac{\hat\xi_{2024} - \hat\xi_{2019}}
     {SE(\hat\xi_{2024} - \hat\xi_{2019})}.
\]

The default reference distribution is standard normal. If you supply a finite
`df`, the function uses a t distribution instead.

## Bootstrap Tests

For objects returned by `bootstrap_popsize()`, the same contrast is evaluated
inside every fractional weighted bootstrap draw. For example, for an increase
test:

\[
d^* = \xi^*_{2024} - \xi^*_{2019}.
\]

The bootstrap standard error is `sd(d*)`, and the function reports a Wald
statistic using that standard error.

```{r bootstrap, eval = requireNamespace("fwb", quietly = TRUE)}
boot_year <- bootstrap_popsize(
  fit,
  R = 19,
  cluster = ~ country_code,
  seed = 2025,
  by = ~ year,
  verbose = FALSE
)

hypotheses_popsize(
  boot_year,
  "xi[year == 2024] - xi[year == 2019] > 0",
  estimate = "boot_median"
)
```

This Wald p value is different from an empirical bootstrap exceedance
probability such as `mean(xi_star > 200000)`. Use
`exceedance_popsize()` for those direct bootstrap tail summaries.

## Power And Limitations

The power of these tests depends on the size of the contrast, the standard
error, the number and quality of observations, clustering or panel dependence,
model specification, and bootstrap stability. A small p value is easier to
obtain when the estimated contrast is large relative to its standard error.

The helper reports the estimate, contrast, standard error, Wald statistic,
p value, S value, and confidence interval. It does not guarantee that a
scientifically important change will be detected with high probability.

Formal power analysis should be simulation-based: choose a data-generating
process, simulate many datasets under the alternative of interest, refit the
model, apply the hypothesis test, and estimate the rejection rate.
