---
title: "iOLS: Iterated OLS for Gamma Pseudo-Maximum Likelihood"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{iOLS: Iterated OLS for Gamma Pseudo-Maximum Likelihood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `uncounted` package supports five estimation methods for the power-law model
$E(m_i) = N_i^{\alpha_i}(\gamma + n_i/N_i)^{\beta_i}$.  This vignette describes
the iOLS (iterated Ordinary Least Squares) estimator, which targets the Gamma
Pseudo-Maximum Likelihood (GPML) solution via a sequence of linear regressions.

The method was proposed by Benatia, Bellego and Pape (2024) as a
computationally simple alternative to nonlinear GLM estimation.  The key
advantage: each iteration is a standard OLS regression, so no gradient or
Hessian computation is needed.

## Model recap

On the log scale the model is linear:

$$\log E(m_i) = \alpha_i \log N_i + \beta_i \log(\gamma + n_i/N_i)
              = \mathbf{z}_i'\boldsymbol{\theta}$$

where $\mathbf{z}_i$ is the design vector and
$\boldsymbol{\theta} = (\alpha, \beta)'$.  This is an exponential conditional
mean model $E(m_i | \mathbf{z}_i) = \exp(\mathbf{z}_i'\boldsymbol{\theta})$.

## GPML vs PPML

There are two leading PML estimators for this model:

**Poisson PML (PPML)** solves $\sum_i (m_i - \mu_i)\mathbf{z}_i = \mathbf{0}$.
It weights observations by $\mu_i$ in the information matrix,
giving more influence to large-count observations.

**Gamma PML (GPML)** solves $\sum_i (m_i/\mu_i - 1)\mathbf{z}_i = \mathbf{0}$.
It gives equal weight to all observations regardless of $\mu_i$, because the
Gamma variance function $V(\mu) = \mu^2$ implies unit working weights under
a log link.

Both are consistent for $\boldsymbol{\theta}$ under correct specification of
the conditional mean.  They differ in efficiency and sensitivity to outliers:
PPML is more efficient when the data are Poisson-like, while GPML is more
robust to observations with unusually large counts relative to $\mu$.

## The iOLS algorithm

iOLS solves the GPML score equations via two phases.

### Phase 1: Warm-up with increasing delta

Initialize $\hat{\boldsymbol{\theta}}_0$ from OLS on $\log(m+1)$.
For each $\delta$ in an increasing sequence (default: 1, 10, 100, 1000):

1. Compute $\hat{\mu}_i = \exp(\mathbf{z}_i'\hat{\boldsymbol{\theta}}_t)$
2. Set $\tilde{y}_i = \log(m_i + \delta\hat{\mu}_i) - \bar{c}$, where
   $\bar{c}$ is the weighted mean of $\log(m_i + \delta\hat{\mu}_i)$
   (empirical centering)
3. Update $\hat{\boldsymbol{\theta}}_{t+1}$ by regressing $\tilde{y}$
   on $\mathbf{Z}$ using OLS
4. Repeat until convergence

This phase is guaranteed to converge globally (contraction mapping with
modulus $\kappa < 1$).  Larger $\delta$ gives a better approximation to
GPML but slower convergence.

### Phase 2: Exact GPML limiting transform

After Phase 1 converges, switch to the exact debiasing step:

$$\tilde{y}_i = \log\hat{\mu}_i + \frac{m_i/\hat{\mu}_i - 1}{1 + \rho}$$

where $\rho > 0$ is a stability parameter (default: 1).  This is the
theoretically correct form from Benatia et al. (2024) that targets the
exact GPML score.  At convergence:

$$\sum_i (m_i/\hat{\mu}_i - 1)\mathbf{z}_i = \mathbf{0}$$

which are the GPML first-order conditions.

## Usage

```{r basic-fit, eval = FALSE}
library(uncounted)
data(irregular_migration)
d <- irregular_migration
d$ukr <- as.integer(d$country_code == "UKR")

# iOLS requires a fixed gamma (estimated gamma not yet supported)
fit_iols <- estimate_hidden_pop(
  data = d, observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "iols",
  cov_alpha = ~ year * ukr,
  cov_beta = ~ year,
  gamma = 0.005,
  countries = ~ country_code
)

summary(fit_iols)
popsize(fit_iols)
```

## Comparison with Poisson

On simulated data with known parameters ($\alpha = 0.70$, $\beta = 0.55$,
$\gamma = 0.005$), all methods recover the truth:

| Method  | $\hat\alpha$ | $\hat\beta$ | $\hat\xi$ |
|---------|-------------|------------|-----------|
| OLS     | 0.713       | 0.583      | 154,609   |
| Poisson | 0.722       | 0.602      | 168,778   |
| NB      | 0.722       | 0.603      | 169,244   |
| iOLS    | 0.714       | 0.587      | 156,714   |

iOLS estimates are closer to OLS than to Poisson/NB, reflecting the
different weighting: GPML gives equal weight to all observations, while
PPML upweights large-$\mu$ observations.

On real migration data with ~50% zeros and complex covariates, iOLS
typically gives **lower** total population estimates than Poisson (about
50--70% of the Poisson total), because GPML downweights the few large
countries that dominate the PPML fit.

## Variance estimation

iOLS uses a sandwich variance estimator with the GPML score residual
$e_i = m_i/\hat{\mu}_i - 1$:

$$\hat{\mathbf{V}} = (\mathbf{Z}'\mathbf{Z})^{-1}
  \mathbf{Z}'\text{diag}(\hat{e}_i^2)\mathbf{Z}
  (\mathbf{Z}'\mathbf{Z})^{-1}$$

The bread $(\mathbf{Z}'\mathbf{Z})^{-1}$ reflects the GPML Fisher
information (unit working weights from the Gamma variance function).
HC0 through HC5 adjustments are available via the `vcov` argument.

Standard errors are generally smaller than Poisson because GPML is more
efficient when the conditional variance is proportional to $\mu^2$
(Gamma-like).  On count data, Poisson SEs may be more conservative.

## Bias correction for population size

The plug-in estimator $\hat{\xi}_g = \sum_{i \in g} N_i^{\hat{\alpha}_g}$ is
upward-biased by Jensen's inequality.  The bias correction uses the
model-based variance $\mathbf{V}_{\text{model}}$, which differs by method:

| Method  | $\mathbf{V}_{\text{model}}$ |
|---------|----------------------------|
| Poisson | $(\mathbf{Z}'\text{diag}(\hat\mu)\mathbf{Z})^{-1}$ |
| iOLS    | $(\mathbf{Z}'\mathbf{Z})^{-1}$ |

For iOLS, the model-based variance is the GPML Fisher information inverse
**without** dispersion scaling.  This is because the Gamma quasi-likelihood
has variance function $V(\mu) = \mu^2$, giving unit working weights.

The current implementation uses a subtractive correction
$\hat{\xi}^{BC} = \hat{\xi} - \text{bias}$, clamped to be positive.
A multiplicative correction based on the exact Gaussian MGF
$\hat{\xi}^{BC} = \sum N_i^{\hat\alpha} \exp(-\frac{1}{2}(\log N_i)^2
\mathbf{x}_i'\hat{\mathbf{V}}\mathbf{x}_i)$ is under development and
would eliminate the negativity issue entirely.

## Convergence diagnostics

Check `fit$convergence`:

- `0`: converged (GPML score condition satisfied)
- `1`: did not converge (consider simpler covariates or checking data)

The normalized GPML score `max|Z'(m/mu - 1)| / n` should be small
(< $10^{-4}$) at convergence.

## Limitations

1. **`gamma = "estimate"` not supported.** The paper's theory covers iOLS
   on the mean model; profiled gamma is a package extension not yet
   validated.  Use a fixed gamma from Poisson estimation.

2. **HC2/HC3 not available for the NB theta-aware path.** This only
   affects NB, not iOLS.

3. **AIC/BIC not comparable across methods.** iOLS uses a GPML pseudo
   log-likelihood, which is not on the same scale as Poisson/NB count
   likelihoods.  `compare_models()` warns about this.

4. **Bootstrap recommended for inference.** The analytical bias correction
   uses the model-based variance, which may underestimate the true
   Jensen's bias on zero-heavy data.  Bootstrap bias correction
   `2*xi_hat - mean(xi_star)` is more robust.

## References

- Benatia, D., Bellego, C. and Pape, L.-D. (2024). Dealing with Logs and
  Zeros in Regression Models. *arXiv preprint* arXiv:2203.11820v3.
- Santos Silva, J. M. C. and Tenreyro, S. (2006). The log of gravity.
  *The Review of Economics and Statistics*, 88(4), 641--658.
- Gourieroux, C., Monfort, A. and Trognon, A. (1984). Pseudo Maximum
  Likelihood Methods: Applications to Poisson Models. *Econometrica*,
  52(3), 701--720.
- Zhang, L.-C. (2008). *Developing methods for determining the number
  of unauthorized foreigners in Norway* (Documents 2008/11). Statistics
  Norway. https://www.ssb.no/a/english/publikasjoner/pdf/doc_200811_en/doc_200811_en.pdf
- Beresewicz, M. and Pawlukiewicz, K. (2020). Estimation of the number
  of irregular foreigners in Poland using non-linear count regression
  models. *arXiv preprint* arXiv:2008.09407.
