---
title: "Model Comparison and Reporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Comparison and Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to compare estimation methods and present
results using the **uncounted** package's plotting and reporting tools.

```{r load, eval = TRUE}
library(uncounted)
data(irregular_migration)
d <- irregular_migration
```

## Fitting multiple models

The package supports five estimation methods. We fit Poisson, NB, and iOLS
to the same data for comparison:

```{r fit-models, eval = TRUE}
fit_po <- estimate_hidden_pop(d, ~ m, ~ n, ~ N,
  method = "poisson", cov_alpha = ~ year + sex,
  gamma = "estimate", countries = ~ country_code)

fit_nb <- estimate_hidden_pop(d, ~ m, ~ n, ~ N,
  method = "nb", cov_alpha = ~ year + sex,
  gamma = fit_po$gamma, countries = ~ country_code)

fit_io <- estimate_hidden_pop(d, ~ m, ~ n, ~ N,
  method = "iols", cov_alpha = ~ year + sex,
  gamma = fit_po$gamma, countries = ~ country_code)
```

## Coefficient tables with modelsummary

The package provides `tidy()` and `glance()` methods compatible with
the **modelsummary** package:

```{r modelsummary, eval = requireNamespace("modelsummary", quietly = TRUE) && requireNamespace("broom", quietly = TRUE)}
library(modelsummary)
modelsummary(
  list(Poisson = fit_po, NB = fit_nb, iOLS = fit_io),
  stars = TRUE
)
```

Individual coefficient tables:

```{r tidy, eval = TRUE}
tidy(fit_po, conf.int = TRUE)
glance(fit_po)
```

## Plotting population size estimates

The `popsize()` function returns an object that can be plotted directly:

```{r plot-popsize, eval = TRUE}
# By year
ps <- popsize(fit_po, by = ~ year)
plot(ps)

# Compare plug-in vs bias-corrected
plot(ps, type = "compare")
```

## Comparing models visually

`compare_popsize()` produces side-by-side population size estimates:

```{r compare-popsize, eval = TRUE}
comp <- compare_popsize(fit_po, fit_nb, fit_io,
  labels = c("Poisson", "NB", "iOLS"),
  by = ~ year)

print(comp)
plot(comp)
```

## Predictions

The `predict()` method supports new data:

```{r predict, eval = TRUE}
# Fitted values
head(predict(fit_po))

# Log-scale (linear predictor)
head(predict(fit_po, type = "link"))

# Predictions for new data
new_d <- d[1:10, ]
predict(fit_po, newdata = new_d)
```

## Model selection guidance

| Situation | Recommended method |
|-----------|-------------------|
| Default analysis | Poisson (robust, well-studied) |
| Overdispersion suspected | NB, then LR test vs Poisson |
| Sensitivity to large countries | iOLS alongside Poisson |
| Quick exploration | OLS with fixed gamma |

When Poisson and iOLS agree, the result is robust to the choice of
weighting. When they diverge, report both and discuss which observations
drive the difference.

## References

- Santos Silva, J. M. C. & Tenreyro, S. (2006). The Log of Gravity.
  *Review of Economics and Statistics*, 88(4), 641--658.
- Benatia, D., Bellego, C. & Pape, L.-D. (2024). Dealing with Logs and
  Zeros in Regression Models. arXiv:2203.11820v3.
- Zhang, L.-C. (2008). *Developing methods for determining the number of
  unauthorized foreigners in Norway* (Documents 2008/11). Statistics Norway.
- Beresewicz, M. & Pawlukiewicz, K. (2020). Estimation of the number of
  irregular foreigners in Poland using non-linear count regression models.
  arXiv:2008.09407.
