---
title: "Theoretical Framework for Hidden Population Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theoretical Framework for Hidden Population Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Estimating the size of unauthorized (hidden) populations is a fundamental
problem in official statistics and migration research.  By definition, these
populations avoid administrative registration, so no single data source
provides a complete count.  However, when multiple *partial* administrative
sources are available -- for example, border-guard apprehensions, police
identifications, and social-insurance records -- statistical models can
exploit cross-source variation to infer the size of the population that
escapes all sources.

The `uncounted` package implements the estimation framework of
Beresewicz and Pawlukiewicz (2020), which builds on the
dual-system ideas of Zhang (2008).  The core insight is that the
*observed* count of unauthorized migrants in a given group (e.g., a
country-of-origin $\times$ year cell) is a power-law function of the
*reference* (known) population, modulated by an auxiliary registration
rate.  Fitting this power-law model and inverting it yields an estimate
of the total hidden population.

This vignette describes the theory behind the model, the available
estimation methods, and the inference procedures implemented in the
package.


## The Model

### Mean structure

Let $i = 1, \ldots, n$ index observational units (e.g., country-of-origin
$\times$ year $\times$ sex cells).  Denote

- $m_i$: the observed count of unauthorized migrants (e.g., apprehensions),
- $N_i$: the reference population size (e.g., insured foreigners from the
  same origin),
- $n_i$: an auxiliary count from a second administrative source (e.g.,
  police identifications).

The conditional mean of $m_i$ is modelled as

$$
E[m_i \mid N_i, n_i] \;=\; \xi_i \rho_i,
  \qquad \xi_i = N_i^{\alpha},
$$

where $\alpha$, $\beta$, and $\gamma \geq 0$ are parameters to be
estimated. Define

$$
\eta_i = \beta \log\!\left(\gamma + \frac{n_i}{N_i}\right).
$$

The package supports three links for $\rho_i = h(\eta_i)$:

- `link_rho = "power"`: $\rho_i = \exp(\eta_i)$,
- `link_rho = "cloglog"`: $\rho_i = 1 - \exp(-\exp(\eta_i))$,
- `link_rho = "logistic"`: $\rho_i = (1 + \exp(-\eta_i))^{-1}$.

On the log scale, this becomes

$$
\log E[m_i] \;=\; \alpha \,\log N_i \;+\; \log \rho_i.
$$

Under every supported detection link,

$$
\xi_i = \frac{E[m_i]}{\rho_i} = N_i^\alpha.
$$

This is why the public population-size estimator remains
$\hat\xi = \sum_i N_i^{\hat\alpha_i}$ even when a bounded detection link is
used. The choice of `link_rho` affects $\hat\xi$ only indirectly through the
fitted value of $\hat\alpha$.

### Parameter interpretation

**$\alpha$ -- elasticity with respect to reference population.**
The parameter $\alpha$ governs how the expected observed count scales with
the reference population.  When $\alpha = 1$, the relationship is
proportional; when $\alpha \in (0, 1)$, the scaling is sublinear, meaning
that larger reference populations yield proportionally fewer detections.
Values of $\alpha$ near zero indicate weak dependence on population size.
The constraint $\alpha \in (0, 1)$ is theoretically motivated but not
always imposed during estimation (see Section "Covariates" below).

**$\beta$ -- elasticity with respect to registration rate.**
The parameter $\beta$ captures the association between the auxiliary
registration rate $(\gamma + n_i / N_i)$ and the observed unauthorized
count.  A positive $\beta$ indicates that groups with higher auxiliary
detection rates also tend to have more observed unauthorized migrants.
This is expected when both sources partially detect the same latent
population.

**$\gamma$ -- offset for zero auxiliary counts.**
Many groups may have $n_i = 0$ (no auxiliary detections), which would
make $\log(n_i / N_i)$ undefined.  The offset $\gamma \geq 0$ ensures
that the rate term remains positive.  In practice, $\gamma$ is typically
small (close to zero).  It can be fixed at a known value, estimated from
the data, or omitted entirely when all $n_i > 0$.

### Population size estimator

Given $\hat\alpha$, the total unauthorized population is estimated as

$$
\hat\xi \;=\; \sum_{i=1}^{n} N_i^{\hat\alpha}.
$$

When $\hat\alpha < 1$ and $N_i > 1$ for all $i$, we have
$\hat\xi < \sum_i N_i$, reflecting the fact that only a fraction of the
reference population is unauthorized.  The population size $\hat\xi$ is
a monotone increasing function of $\alpha$ (for $N_i \geq 1$), a
property exploited for confidence interval construction.

```{r basic-model, eval = TRUE}
library(uncounted)
data(irregular_migration)

## Basic Poisson model with estimated gamma
fit <- estimate_hidden_pop(
  data      = irregular_migration,
  observed  = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method    = "poisson"
)
summary(fit)

## Population size estimate with 95% CI
popsize(fit)
```


## Covariates

### Covariate-varying parameters

The scalar parameters $\alpha$ and $\beta$ can be made group-specific
through design matrices.  Let $\mathbf{X}_{\alpha,i}$ and
$\mathbf{X}_{\beta,i}$ be covariate row vectors for observation $i$.
The unconstrained specification is

$$
\alpha_i = \mathbf{X}_{\alpha,i}^\top \mathbf{a}, \qquad
\beta_i  = \mathbf{X}_{\beta,i}^\top \mathbf{b},
$$

where $\mathbf{a}$ and $\mathbf{b}$ are coefficient vectors.

### Constrained estimation

In the unconstrained case, nothing prevents $\hat\alpha_i$ from falling
outside $(0, 1)$ or $\hat\beta_i$ from being negative.  When theoretical
constraints are desired, link functions can be applied:

$$
\alpha_i = \mathrm{logit^{-1}}(\mathbf{X}_{\alpha,i}^\top \mathbf{a})
  = \frac{1}{1 + \exp(-\mathbf{X}_{\alpha,i}^\top \mathbf{a})},
  \qquad
\beta_i = \exp(\mathbf{X}_{\beta,i}^\top \mathbf{b}),
$$

ensuring $\alpha_i \in (0, 1)$ and $\beta_i > 0$.  Constrained
estimation is available for the Poisson and Negative Binomial methods
(set `constrained = TRUE`).  Coefficients are then reported on the link
scale; response-scale values are shown by `summary()`.

```{r covariates, eval = TRUE}
## Alpha and beta varying by sex
fit_cov <- estimate_hidden_pop(
  data      = irregular_migration,
  observed  = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method    = "poisson",
  cov_alpha = ~ sex,
  cov_beta  = ~ sex
)
summary(fit_cov)

## Constrained Poisson: alpha in (0,1), beta > 0
fit_constr <- estimate_hidden_pop(
  data      = irregular_migration,
  observed  = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method    = "poisson",
  cov_alpha = ~ sex,
  constrained = TRUE
)
summary(fit_constr)
```


## Estimation Methods

The `uncounted` package implements four estimation methods.  All share
the same mean structure; they differ in the objective function and
distributional assumptions.

### OLS on the log scale

Ordinary least squares applied to the log-linearized model:

$$
\log m_i = \alpha \log N_i + \beta \log\!\left(\frac{n_i + \gamma}{N_i}\right) + \varepsilon_i.
$$

When some $m_i = 0$, the response is replaced by $\log(m_i + 1)$.  OLS
is fast and transparent but ignores the count nature of $m_i$ and may be
biased under heteroscedasticity (the log-transformation changes the
conditional mean).

When $\gamma$ is estimated, it is profiled out via grid search: for each
candidate $\gamma$, the OLS coefficients are obtained in closed form, and
the value minimizing the residual sum of squares is selected.

### NLS on the original scale

Nonlinear least squares minimizes

$$
\sum_{i=1}^n \left(m_i - N_i^{\alpha}
  \left(\frac{n_i + \gamma}{N_i}\right)^{\!\beta}\right)^2
$$

using the Levenberg--Marquardt algorithm.  This avoids the
log-transformation bias of OLS but still treats $m_i$ as continuous.

### Poisson pseudo-maximum likelihood (PPML)

The Poisson log-likelihood is

$$
\ell(\boldsymbol{\theta}) = \sum_{i=1}^{n}
  \bigl[m_i \log \mu_i - \mu_i\bigr],
$$

where $\mu_i = N_i^{\alpha_i}\,(\gamma + n_i/N_i)^{\beta_i}$.  As shown by
Santos Silva and Tenreyro (2006), the Poisson PML estimator is consistent
for the parameters of the conditional mean $E[m_i \mid N_i, n_i]$ even if
the true distribution is not Poisson, provided the mean is correctly
specified.  This makes PPML robust to heteroscedasticity and is the
recommended default in `uncounted`.

For the Poisson and Negative Binomial count models, the package can also use
`estimator = "gmm"` or `estimator = "el"`. In those cases the same
per-observation score equations are treated as moment conditions and fitted
with `momentfit`. The model for $\mu_i$ is unchanged; only the estimating
criterion differs. For these moment-based count fits the default robust
covariance is HC1 rather than HC3, because leverage-based HC2/HC3
adjustments are not available generically in that setting.

### Negative Binomial ML

The Negative Binomial extends the Poisson by adding a dispersion parameter
$\theta > 0$.  The log-likelihood is

$$
\ell(\boldsymbol{\theta}, \theta) = \sum_{i=1}^{n} \Bigl[
  \log\Gamma(m_i + \theta) - \log\Gamma(\theta) - \log(m_i!)
  + \theta\log\theta - \theta\log(\theta + \mu_i)
  + m_i\log\mu_i - m_i\log(\theta + \mu_i)
\Bigr].
$$

This accounts for overdispersion (variance exceeding the Poisson mean)
and can provide more efficient estimates when the data are truly
overdispersed.  The parameter $\theta$ is jointly estimated alongside
$\alpha$, $\beta$, and $\gamma$.

### iOLS (Gamma PML)

The iOLS estimator (Benatia, Bellego & Pape, 2024) targets the Gamma
Pseudo-Maximum Likelihood (GPML) score equations

$$\sum_i (m_i / \mu_i - 1)\,\mathbf{z}_i = \mathbf{0}$$

via a two-phase iterated OLS algorithm.  Unlike PPML, which weights
observations by $\mu_i$, GPML gives equal weight to all observations.
This makes it more robust to influential high-count observations.
See `vignette("iols")` for the full algorithm and comparison.

Currently `gamma = "estimate"` is not supported for iOLS; use a fixed
gamma from a Poisson fit.

### Gamma estimation

The offset parameter $\gamma$ is handled differently depending on the
method:

- **Poisson / NB:** $\gamma$ is jointly optimized within the likelihood
  using L-BFGS-B with box constraints (`gamma_bounds`, default
  $[10^{-10},\, 0.5]$).
- **OLS:** $\gamma$ is profiled out via a grid search over the bounds,
  followed by local optimization (`optimize()`).
- **All methods:** $\gamma$ can be fixed at a known value (`gamma = 0.01`)
  or omitted entirely (`gamma = NULL`), the latter requiring $n_i > 0$ for
  all observations.

```{r methods, eval = TRUE, warning = FALSE}
## Compare estimation methods
fit_ols <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "ols"
)

fit_pois <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "poisson"
)

fit_nb <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "nb"
)

fit_iols <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m, auxiliary = ~ n, reference_pop = ~ N,
  method = "iols", gamma = fit_pois$gamma
)

## Extract population size estimates
popsize(fit_ols)
popsize(fit_pois)
popsize(fit_nb)
popsize(fit_iols)
```


## Inference

### Robust standard errors

All estimation methods report heteroscedasticity-consistent (HC) robust
standard errors computed via the sandwich estimator.  The general form is

$$
\widehat{\mathrm{Var}}(\hat{\boldsymbol{\theta}}) =
  (\mathbf{J}^\top \mathbf{J})^{-1}\,
  \mathbf{J}^\top \,\mathrm{diag}(\tilde{e}_1^2, \ldots, \tilde{e}_n^2)\,
  \mathbf{J}\,
  (\mathbf{J}^\top \mathbf{J})^{-1},
$$

where $\mathbf{J}$ is the Jacobian (or score matrix) and $\tilde{e}_i$
are adjusted residuals.  The adjustment depends on the HC variant:

| Variant | Residual adjustment |
|---------|---------------------|
| HC0     | $\tilde{e}_i = e_i$ (no correction) |
| HC1     | $\tilde{e}_i = e_i \sqrt{n/(n-p)}$ |
| HC2     | $\tilde{e}_i = e_i / \sqrt{1 - h_{ii}}$ |
| HC3     | $\tilde{e}_i = e_i / (1 - h_{ii})$ (default) |
| HC4     | $\tilde{e}_i = e_i / (1 - h_{ii})^{\delta_i/2}$ |
| HC5     | $\tilde{e}_i = e_i / (1 - h_{ii})^{\min(\delta_i, h_{\max})/2}$ |

Here $h_{ii}$ are the hat-matrix diagonal elements (leverages) and
$\delta_i$ is a data-dependent exponent. HC3 is the default for OLS,
NLS, Poisson MLE, NB MLE, and iOLS, and is recommended for moderate
sample sizes as it provides a good balance between finite-sample bias
correction and stability. For the moment-based count estimators
(`estimator = "gmm"` and `estimator = "el"`), the default robust
covariance is HC1.

### Delta-method confidence intervals for population size

The population size $\hat\xi = \sum_i N_i^{\hat\alpha}$ is a nonlinear
function of $\hat\alpha$. The implemented subgroup intervals are based on
the delta method for $\hat\xi_g$ itself, followed by a log-normal
approximation to preserve positivity.

For a subgroup $g$, let $\mathbf{g}_g$ denote the gradient of
$\hat\xi_g = \sum_{i \in g} N_i^{\hat\alpha_i}$ with respect to the alpha
coefficients. In the unconstrained case,

$$
\mathbf{g}_g =
  \sum_{i \in g} N_i^{\hat\alpha_i}\log(N_i)\mathbf{x}_i.
$$

When `constrained = TRUE`, the same expression is multiplied by the
derivative of the inverse-logit map,
$\hat\alpha_i(1-\hat\alpha_i)$, because the model is parameterized on the
link scale.

Using the HC-robust covariance $\mathbf{V}$ for the alpha coefficients,

$$
\widehat{\mathrm{Var}}(\hat\xi_g) =
  \mathbf{g}_g^\top \mathbf{V} \mathbf{g}_g.
$$

The subgroup standard error
$\widehat{\mathrm{se}}(\hat\xi_g)=\sqrt{\widehat{\mathrm{Var}}(\hat\xi_g)}$
is then mapped to a log-normal interval

$$
\left(
\hat\xi_g \exp\!\left\{-z_{\alpha/2}
  \frac{\widehat{\mathrm{se}}(\hat\xi_g)}{\hat\xi_g}\right\},
\;
\hat\xi_g \exp\!\left\{z_{\alpha/2}
  \frac{\widehat{\mathrm{se}}(\hat\xi_g)}{\hat\xi_g}\right\}
\right).
$$

When analytical bias correction is requested, the package rescales both
bounds by the ratio $\hat\xi_g^{BC}/\hat\xi_g$ so that the returned
interval is centered on the bias-corrected estimate.

**Total across groups.**  When the model includes covariates in $\alpha$,
the total $\hat\xi = \sum_g \hat\xi_g$ has its own delta-method CI
computed via the gradient $\nabla_\alpha \xi$ and a log-normal
approximation to ensure positivity.

### Bias correction

Because $\xi(\alpha) = \sum_i N_i^\alpha$ is convex in $\alpha$ for
$N_i > 1$, Jensen's inequality implies $E[\hat\xi] \geq \xi$. Writing
$N_i^{\hat\alpha_i} = \exp(\hat\alpha_i \log N_i)$ and treating
$\hat{\boldsymbol{\alpha}}$ as approximately normal gives

$$
E[N_i^{\hat\alpha_i}] \approx N_i^{\alpha_i}
  \exp\!\left\{\frac{1}{2}(\log N_i)^2
  \mathbf{x}_i^\top \mathbf{V}_{\mathrm{model}} \mathbf{x}_i\right\}.
$$

Inverting this yields the multiplicative lognormal correction used by
the package for unconstrained models:

$$
\hat\xi^{BC} =
  \sum_{i=1}^{n}
  N_i^{\hat\alpha_i}
  \exp\!\left(
    -\frac{1}{2}(\log N_i)^2
    \mathbf{x}_i^\top \mathbf{V}_{\mathrm{model}} \mathbf{x}_i
  \right).
$$

This correction is strictly positive and exact under the Gaussian
approximation for the linear predictor. For constrained fits, where
$\alpha_i = \operatorname{logit}^{-1}(\eta_i)$, the corresponding
logistic-normal integral has no closed form, so the package falls back
to a second-order Taylor approximation on the link scale:

$$
\mathrm{Bias}(\hat\xi) \;\approx\;
  \frac{1}{2}\,\sum_{i=1}^{n} N_i^{\alpha}\,(\log N_i)^2 \;\cdot\;
  \mathbf{x}^\top \mathbf{V}_{\mathrm{model}}\,\mathbf{x},
$$

where $\mathbf{V}_{\mathrm{model}}$ is the model-based (homoscedastic)
variance--covariance matrix.

Model-based variance is used rather than the HC-robust variance, because
the latter can be inflated by high-leverage observations in skewed data,
leading to overcorrection. For Poisson and NB fits, `vcov_model` is the
Fisher-style inverse information for the mean-model parameters. This
remains true when the coefficients are obtained by `estimator = "gmm"`
or `estimator = "el"`: the package evaluates the same Fisher-style model
variance at the non-MLE estimate and reserves the robust HC or FWB
covariance for interval estimation. Bias correction is also applied to
the CI bounds, evaluated at their respective $\alpha$ values.

### Fractional weighted bootstrap

For finite-sample inference and as a robustness check on the analytical
CIs, the package provides the fractional weighted bootstrap (FWB) of
Xu et al. (2020) via `bootstrap_popsize()`.

Instead of resampling rows, FWB generates random non-negative weights
$w_1, \ldots, w_n$ from a $\mathrm{Dirichlet}(1, \ldots, 1)$
distribution and refits the model under these weights.  This avoids
duplicate-row issues with discrete data and is naturally suited to
weighted GLMs.

**Cluster bootstrap.**  When observations are not independent (e.g., the
same country observed across years and sexes), a cluster variable can be
specified.  In this case, all observations within the same cluster
receive the same FWB weight, preserving within-cluster correlation.

**Bootstrap CIs.**  Two types are available:

- *Percentile*: the $\alpha/2$ and $(1 - \alpha/2)$ quantiles of the
  bootstrap distribution.
- *Bias-corrected percentile (BC)*: adjusts the quantile probabilities
  for median bias using the standard BC formula.

The bootstrap median is the default point estimate because the
transformation $\xi = \sum N_i^\alpha$ is convex, so the bootstrap
distribution of $\hat\xi$ is right-skewed and the bootstrap mean
tends to overestimate $\xi$.

```{r bootstrap, eval = requireNamespace("fwb", quietly = TRUE)}
## Fit a model with country-level grouping
fit <- estimate_hidden_pop(
  data      = irregular_migration,
  observed  = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method    = "poisson",
  countries = ~ country
)

## Cluster bootstrap with 199 replicates
boot_result <- bootstrap_popsize(
  fit,
  R       = 49,
  cluster = ~ country_code,
  seed    = 42
)
boot_result

## Access the full results table
boot_result$popsize_full
```


## References

- Beręsewicz, M., & Pawlukiewicz, K. (2020). Estimation of the number of irregular foreigners in Poland using non-linear count regression models. arXiv preprint arXiv:2008.09407.

- Santos Silva, J. M. C. and Tenreyro, S. (2006). The log of gravity.
  *The Review of Economics and Statistics*, 88(4), 641--658.

- Xu, L., Gotwalt, C., Hong, Y., King, C. B., and Meeker, W. Q. (2020).
  Applications of the fractional-random-weight bootstrap. *The American
  Statistician*, 74(4), 345--358.

- Zhang, L.-C. (2008). *Developing methods for determining the number of unauthorized foreigners in Norway* (Documents 2008/11). Statistics Norway. https://www.ssb.no/a/english/publikasjoner/pdf/doc_200811_en/doc_200811_en.pdf
- Beręsewicz, M., & Pawlukiewicz, K. (2020). Estimation of the number of irregular foreigners in Poland using non-linear count regression models. arXiv preprint arXiv:2008.09407.
