---
title: "Complete Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette walks through a complete analysis estimating the unauthorized
foreign population in Poland using the `irregular_migration` dataset shipped
with the **uncounted** package. The workflow covers every step from loading the
data through fitting models, obtaining population size estimates with confidence
intervals, running bootstrap inference, selecting among competing specifications,
and performing sensitivity analysis. By the end you will have a reproducible
pipeline that can be adapted to other datasets.

```{r load-package}
library(uncounted)
```

## Data Exploration

The `irregular_migration` dataset is a country-level panel (2019--2024) with
three count variables per observation:

- **m** -- foreigners apprehended by the Border Guard for unauthorized stay
  (the quantity whose hidden population we want to estimate).
- **n** -- foreigners identified by the Police (an auxiliary, partially
  overlapping administrative source).
- **N** -- foreigners registered in the Social Insurance Institution (ZUS),
  used as a proxy for the total known foreign population from a given country.

```{r data-overview}
data(irregular_migration)
str(irregular_migration)
```

A large share of observations have zero apprehensions, reflecting the many small
countries of origin with no recorded unauthorized stay.

```{r zero-shares}
cat("Share m == 0:", round(mean(irregular_migration$m == 0) * 100, 1), "%\n")
cat("Share n == 0:", round(mean(irregular_migration$n == 0) * 100, 1), "%\n")
```

A log--log scatter of the apprehension count against the reference population
reveals the power-law relationship that motivates the model. Observations with
`m > 0` and `n > 0` are shown.

```{r explore-plot, fig.width = 7, fig.height = 4}
pos <- irregular_migration$m > 0 & irregular_migration$n > 0
plot(
  log(irregular_migration$N[pos]),
  log(irregular_migration$m[pos]),
  xlab = "log(N)", ylab = "log(m)",
  main = "Apprehensions vs reference population (log-log)",
  pch = 16, cex = 0.6, col = adjustcolor("steelblue", 0.6)
)
```

## Fitting Models

We fit four specifications, all using the same formula interface. The core
model is

$$E(m_i) = N_i^{\alpha}\,(\gamma + n_i / N_i)^{\beta},$$

where $\alpha$ governs the elasticity with respect to the reference population,
$\beta$ captures the relationship with the auxiliary detection rate, and $\gamma$
is a baseline offset ensuring the rate term is positive even when $n_i = 0$.

### Model 1: Poisson (unconstrained, estimated gamma)

This is the recommended default. Poisson pseudo-maximum likelihood (PPML) is
consistent for the conditional mean even under heteroscedasticity, and gamma is
estimated jointly.

```{r fit-poisson}
fit_pois <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method = "poisson",
  gamma = "estimate",
  countries = ~ country
)
summary(fit_pois)
```

Key output to look for:

- **alpha** -- the elasticity with respect to `N`. Values below 1 indicate
  that the unauthorized population grows less than proportionally with the
  reference population.
- **beta** -- the elasticity with respect to the detection rate. Positive
  values mean higher auxiliary rates are associated with more apprehensions.
- **gamma** -- the estimated baseline offset (typically small).

### Model 2: Poisson constrained (alpha in (0,1), beta > 0)

Constraining alpha to the unit interval and beta to be positive enforces
theoretical expectations. With `constrained = TRUE`, the fitted
observation-level parameters are
`alpha_values = plogis(X_alpha %*% alpha_coefs)` and
`beta_values = exp(X_beta %*% beta_coefs)`, so fitted alpha stays in
`(0, 1)` and fitted beta stays positive. The reported coefficient vectors
remain on the transformed scales (logit for alpha, log for beta).

```{r fit-poisson-constr}
fit_pois_c <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method = "poisson",
  gamma = "estimate",
  constrained = TRUE,
  countries = ~ country
)
summary(fit_pois_c)
```

The `summary()` output reports both link-scale coefficients (with standard
errors and p-values) and response-scale alpha/beta values with delta-method
standard errors.

### Model 3: Negative Binomial (estimated gamma)

The NB model adds a dispersion parameter theta to accommodate overdispersion
beyond what Poisson allows.

```{r fit-nb}
fit_nb <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method = "nb",
  gamma = "estimate",
  countries = ~ country
)
summary(fit_nb)
```

### Model 4: NB two-stage (gamma from Poisson, fixed in NB)

A pragmatic two-stage approach: first estimate gamma from the Poisson model
(which is better identified), then fix it in the NB model. This avoids
potential numerical instability when estimating gamma and theta simultaneously.

```{r fit-nb-twostage}
gamma_from_pois <- fit_pois$gamma

fit_nb2 <- estimate_hidden_pop(
  data = irregular_migration,
  observed = ~ m,
  auxiliary = ~ n,
  reference_pop = ~ N,
  method = "nb",
  gamma = gamma_from_pois,
  countries = ~ country
)
summary(fit_nb2)
```

## Population Size Estimates

The `popsize()` function extracts the estimated total unauthorized population
$\hat{\xi} = \sum_i N_i^{\hat{\alpha}}$ along with bias-corrected estimates
and confidence intervals.

```{r popsize-pois}
ps_pois <- popsize(fit_pois)
ps_pois
```

The columns are:

- **group** -- label for each alpha-covariate group (here `"(all)"` since we
  have no covariates on alpha).
- **observed** -- total observed count $\sum m_i$.
- **estimate** -- plug-in estimate $\hat{\xi}$.
- **estimate_bc** -- analytical bias-corrected estimate. For unconstrained
  fits this is the multiplicative lognormal correction; constrained fits use
  a second-order Taylor approximation.
- **lower**, **upper** -- confidence interval bounds. The CI is constructed
  by a delta-method/log-normal approximation on the population-size scale.
  When bias correction is requested, the bounds are rescaled by the same
  `estimate_bc / estimate` ratio as the reported point estimate.
- **share_pct** -- group share as a percentage of the total.

Comparing across all four models:

```{r popsize-compare}
models <- list(
  Poisson       = fit_pois,
  Poisson_constr = fit_pois_c,
  NB            = fit_nb,
  NB_twostage   = fit_nb2
)

ps_table <- do.call(rbind, lapply(names(models), function(nm) {
  ps <- popsize(models[[nm]])
  data.frame(
    Model    = nm,
    Estimate = sum(ps$estimate),
    BC       = sum(ps$estimate_bc),
    Lower    = sum(ps$lower),
    Upper    = sum(ps$upper)
  )
}))
ps_table
```

### Stratified population size

The `by` parameter allows computing population size by any variable in the
data, not just the `cov_alpha` groups. This is useful when alpha varies by
sex but you want totals by year, continent, or country.

```{r popsize-by-year, eval = TRUE}
# Population size by year (using the same fitted alpha)
popsize(fit_pois, by = ~ factor(year))

# By continent
popsize(fit_pois, by = ~ continent)

# By country (one estimate per country)
popsize(fit_pois, by = ~ country)
```

The total across all `by`-groups equals the total from the default grouping.

## Bootstrap Inference

The `bootstrap_popsize()` function uses the fractional weighted bootstrap
(FWB) of Xu et al. (2020). Cluster bootstrap by country is recommended
to account for within-country correlation across years.

```{r bootstrap, eval = requireNamespace("fwb", quietly = TRUE)}
set.seed(2025)
boot_pois <- bootstrap_popsize(
  fit_pois,
  R = 49,
  cluster = ~ country_code,
  ci_type = "perc",
  seed = 2025
)
boot_pois
# #> Bootstrap population size estimation
# #> R = 49 | CI type: perc | Point estimate: median | Converged: 998 / 999
# #> Cluster bootstrap
# #> 95% CI
```

Access the detailed results:

```{r bootstrap-details, eval = requireNamespace("fwb", quietly = TRUE)}
# Main table (selected point estimate + CI)
boot_pois$popsize

# Full table with all point estimate types
boot_pois$popsize_full

# Bootstrap distribution summary
summary(boot_pois)
```

### Comparing bootstrap and analytical CIs

```{r ci-compare, eval = TRUE}
ps_analytical <- popsize(fit_pois)
data.frame(
  Type       = c("Analytical", "Bootstrap"),
  Estimate   = c(sum(ps_analytical$estimate_bc),
                 boot_pois$popsize$estimate),
  Lower      = c(sum(ps_analytical$lower),
                 boot_pois$popsize$lower),
  Upper      = c(sum(ps_analytical$upper),
                 boot_pois$popsize$upper)
)
```

The bootstrap median is typically smaller than the plug-in estimate. This is
expected: $\xi(\alpha) = \sum N_i^\alpha$ is convex in $\alpha$ for $N_i > 1$,
so by Jensen's inequality $E[\hat{\xi}] \geq \xi$. The bootstrap mean inherits
this upward bias, while the bootstrap median is more robust to the right skew
of the bootstrap distribution and is therefore the recommended point estimate.

## Model Selection

### Information criteria and fit statistics

```{r compare-models}
comp <- compare_models(
  Poisson        = fit_pois,
  Poisson_constr = fit_pois_c,
  NB             = fit_nb,
  NB_twostage    = fit_nb2,
  sort_by = "AIC"
)
comp
```

The output includes three pseudo $R^2$ measures: `R2_cor` (squared
correlation), `R2_D` (explained deviance relative to a null model without
covariates, see Cameron & Trivedi, 2013), and `R2_CW` (Cameron--Windmeijer,
1997, which accounts for the model's variance function).

**Note on `R2_D`**: The null model is the same specification (same method,
same gamma) but without `cov_alpha` / `cov_beta` --- i.e., a single $\alpha$
and single $\beta$. If your model already has no covariates, then
model = null and `R2_D = 0` by construction. `R2_D` becomes informative when
you compare models with covariates (e.g., `cov_alpha = ~ year + sex`) against
the covariate-free baseline: it tells you how much of the deviance is
explained by group-varying parameters beyond the basic power-law structure.

### Likelihood ratio test: Poisson vs NB

```{r lrtest}
lr <- lrtest(fit_pois, fit_nb)
lr
```

When comparing Poisson (H0) against NB (H1), the dispersion parameter theta
lies on the boundary of its parameter space under H0 ($\theta \to \infty$).
The function automatically applies the Self & Liang (1987) correction,
yielding $p = 0.5 \cdot \Pr(\chi^2_1 > LR)$.

### Diagnostics

The four-panel diagnostic plot shows fitted vs observed, Anscombe residuals,
scale-location, and a normal Q-Q plot.

```{r diagnostics-pois, fig.width = 7, fig.height = 7}
par(mfrow = c(2, 2))
plot(fit_pois, ask = FALSE)
```

The rootogram compares observed and fitted count frequencies. Hanging bars
that touch the zero line indicate a good fit.

```{r rootogram-pois, fig.height = 4}
rootogram(fit_pois)
```

```{r rootogram-nb, fig.height = 4}
rootogram(fit_nb)
```

### Why Poisson as the main model?

Despite the NB often having a better AIC/BIC (reflecting genuine
overdispersion), the Poisson model is recommended as the primary specification
for several reasons:

1. **Consistency.** Poisson PML is consistent for the conditional mean as long
   as $E(m_i | X_i)$ is correctly specified, regardless of the true variance
   function. The NB requires correct specification of both the mean and the
   variance.
2. **Robustness.** HC-robust standard errors correct for overdispersion in
   the Poisson model without adding an extra parameter that must be estimated.
3. **Stability.** The NB model's joint estimation of gamma and theta can be
   numerically fragile when many observations are zero. The two-stage approach
   mitigates this but introduces a conditioning step.

The NB results serve as a useful robustness check on the population size
estimate.

### Exploratory log-log plots

The `plot_explore()` function produces marginal log-log scatterplots that
visualise the power-law relationships.

```{r explore-fit, fig.width = 7, fig.height = 4}
plot_explore(fit_pois)
```

## Sensitivity Analysis

### Leave-one-out by country

LOO analysis refits the model dropping one country at a time and reports the
change in $\hat{\xi}$.

```{r loo-country, eval = TRUE}
loo_pois <- loo(fit_pois, by = "country")
print(loo_pois)
# #> Leave-one-out sensitivity analysis
# #> Dropped by: country
# #> ...
# #> Most influential (by |delta xi|):
# #>   dropped      dxi  pct_change
# #>   Ukraine  -12345.0       -8.50
# #>   ...
```

```{r loo-plot, eval = TRUE}
plot(loo_pois)
```

The coefficient stability summary shows how much alpha and beta shift:

```{r loo-summary, eval = TRUE}
summary(loo_pois)
```

### Comparing LOO across models

```{r compare-loo, eval = TRUE}
loo_nb <- loo(fit_nb, by = "country")
comp_loo <- compare_loo(loo_pois, loo_nb, labels = c("Poisson", "NB"))
print(comp_loo)
plot(comp_loo, type = "scatter")
plot(comp_loo, type = "bar")
```

The scatter plot shows whether the same countries are influential under both
models. Points near the diagonal indicate agreement; points in off-diagonal
quadrants indicate divergent influence.

### Gamma profile

The `profile_gamma()` function refits the model across a grid of fixed gamma
values and shows how $\hat{\xi}$ and the log-likelihood depend on gamma. A
flat xi profile indicates robustness; a steep profile suggests sensitivity.

```{r gamma-profile, fig.width = 7, fig.height = 4}
prof <- profile_gamma(fit_pois)
```

The returned data frame can be inspected directly:

```{r gamma-profile-data}
head(prof)
```

## Reporting Results

To prepare results for a paper, combine the analytical and bootstrap estimates
into a single summary table.

```{r results-table, eval = TRUE}
# Analytical results
ps_a <- popsize(fit_pois)

# Bootstrap results (assuming boot_pois was computed above)
ps_b <- boot_pois$popsize_full

results <- data.frame(
  Observed         = sum(ps_a$observed),
  Plugin           = round(sum(ps_a$estimate)),
  Plugin_BC        = round(sum(ps_a$estimate_bc)),
  Analytical_Lower = round(sum(ps_a$lower)),
  Analytical_Upper = round(sum(ps_a$upper)),
  Boot_Median      = round(sum(ps_b$boot_median)),
  Boot_Lower       = round(sum(ps_b$lower)),
  Boot_Upper       = round(sum(ps_b$upper))
)
results
```

Extract coefficients for reporting:

```{r coef-table}
coefs <- coef(fit_pois)
se <- sqrt(diag(vcov(fit_pois)))
ci_lo <- coefs - 1.96 * se
ci_hi <- coefs + 1.96 * se

coef_table <- data.frame(
  Parameter = names(coefs),
  Estimate  = round(coefs, 4),
  SE        = round(se, 4),
  CI_lower  = round(ci_lo, 4),
  CI_upper  = round(ci_hi, 4),
  row.names = NULL
)
coef_table
```

The estimated gamma and model fit statistics:

```{r fit-stats}
cat("Gamma:", round(fit_pois$gamma, 6), "\n")
cat("Log-likelihood:", round(fit_pois$loglik, 2), "\n")
cat("AIC:", round(AIC(fit_pois), 2), "\n")
cat("BIC:", round(BIC(fit_pois), 2), "\n")
cat("Deviance:", round(deviance(fit_pois), 2), "\n")
```
